---
title: "PS270 Final Project"
author: "Samantha Dombar"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

```{r}
install.packages("distill", repos = "https://cloud.r-project.org")
library(distill)
```

## Project thoughts

I am interested in exploring data related to social media and deomocracy, whether that be advertising statistics or social media and self identification.


## Data Set & Research Question

The research question I will be answering using regression analysis is: Do wartime protests in democratic Middle Eastern or North African Countries cause citizens to vote out incumbent leaders?

Using data from the Varieties of Democracy Project (V-DEM), I will examine how protest mobilization relates to political outcomes in the Middle East and North Africa (MENA) region. Specifically, I will draw on V-DEM’s regional classification variable identifying countries in the MENA region (Region #3) and the dataset’s protest mobilization indicator, which captures the extent to which citizens engage in organized demonstrations or collective action against political actors. The independent variable will be the level of protest mobilization within a country-year, as measured by the V-DEM protest mobilization variable. The dependent variable will be the political response to this mobilization, such as shifts in electoral outcomes, government support, or opposition strength within the region. I hypothesize that moderate levels of protest mobilization will increase the likelihood that citizens oppose incumbent leaders or ruling parties, as protests may signal dissatisfaction and mobilize broader political participation. However, extremely high levels of protest activity may produce the opposite effect if citizens become discouraged by instability, fear repression, or rally behind incumbents in response to perceived disorder. A positive and statistically significant relationship between protest mobilization and opposition support would support my hypothesis. Conversely, a statistically significant negative relationship would suggest that protest activity either fails to translate into opposition gains or may even strengthen incumbent support under certain conditions.


## Set-up code

```{r}
devtools::install_github("vdeminstitute/vdemdata")
library(vdemdata)
vdem <- vdemdata::vdem
head(vdem)
```
## Data Cleaning

```{r}
mena <- c("Algeria","Bahrain","Egypt","Iran","Iraq","Israel",
          "Jordan","Kuwait","Lebanon","Libya","Morocco",
          "Oman","Qatar","Saudi Arabia","Syria","Tunisia",
          "United Arab Emirates","Yemen")

```

```{r}
vdem_clean <- vdem |>
  dplyr::filter(country_name %in% mena) |>
  dplyr::filter(v2ellostss != 1) |>
  dplyr::select(
    country_name, country_text_id, year, v2x_polyarchy, v2x_libdem,v2csantimv,
    v2ellostss)

vdem_clean
```

## Graphing

The graph below shows the distribution of anti government movements in Middle Eastern and North African Countries in the dataset.

```{r}
boxplot(v2csantimv ~ country_name,
        data = vdem_clean,
        las = 2,
        main = "Distribution of Anti-Government Movements by MENA Country",
        xlab= "Country Name",
        ylab = "Anti-Government Movements")
```


